Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-24 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
changes, mostly small

The power of the label? Getting the word out about RSE support using the label. 
\choinote{Please review this short, fragmented paragraph.}

@@

An alternative approach has been the provision of programming expertise as part
of national computing services. The US XSEDE project's Extended Collaboration Support 
Services (ECSS) is a set of developers who are paid with XSEDE funding, and are
on ``permanent'' staff. When PIs request allocations on XSEDE resources, there
is a finite pool of developer time that can be awarded, typically for one year
only, and at partial effort, typically 20 percent or so. The finite time allowed
provides motivation for the scientists and their group to work closely with a
developer and to become educated in what the developer is doing, so they can
sustain the effort once the ECSS period is over. This funding mechanism can be
highly efficient for scientific problems, because the developer pool assembled
by the research providers are, by definition, expert in the characteristics of
their specific resource, and can very quickly assess the scientists' needs, and
what it will take to implement software that meets the user's needs. However, it
does not develop capacity within institutions, and since XSEDE is a time-bounded
program, it should not be relied upon as a long-term solution to acquiring this
type of capacity.

The UK allows this kind of collaboration, called ARCHER, to support the creation of scientific
software for the large supercomputing resource. However, as well
as resourcing this from the staff of the Edinburgh Parallel Computing Centre, \choinote{not sure what the previous part means}
who host the computer, this ``embedded CSE'' resource also allows the
programming to come from local groups. This has been very helpful in providing
funding to establish local groups. These work best when they develop good
collaborations with national cyberinfrastructure pools. When an organization
assembles a developer pool, diversity is developed and skills can be
transferred.

@@

The CANARIE group (Canada) accepts proposals for providing services to broad
communities, integrating people who are doing things that are complementary. The
goal is to make the available stack more robust and richer for everyone. They
offer short rounds of funding that can have as a key metric creating some useful
functionality that shows a diversity of input and draws from across disciplines,
then more funding could follow. \choinote{I find the last sentence unclear.} This could apply within or across institutions.

There can be communication problems across cultural barriers, with domain
scientists seeing developers as ``other''. Collaboration, and tools to fund,
encourage or motivate collaboration are extremely important. 
\choinote{Last sentence have at least two verbs.}

We think support from non-governmental organizations will be important if RSE
groups will become established. The Sloan Foundation is currently funding data
science engineers, who work in the context of other SW developers at the University
of Washington. These scientists work in the e-Science Studio/Data Science
Studio, and they help a group of graduate students in solving their problems in
data science and data management. During Fall and Spring, a 10-week incubator
program allows students to work two days a week on a data-intensive
science project. Some fraction of the developer time is dedicated to the
developers' personal interests as well as instructions.

The goal for Sloan is to provide success stories and demonstrable value
in the presence of data scientists on university staff. These stories are the
basis for arguments to the host organization. This is an effort to create
awareness of the value of research scientist developers. Embedding with
scientists, and adding spare capacity is critical to making the innovation
possible. This model is essentially to argue for permanent budget lines to
support data scientists as part of university staff hires, just as with core
facilities. This could become a fee-for-service model requested by grant
funding, just as DNA sequencing is for core facilities, if it becomes apparent
that this gives competitive advantages to a university's research effort.

One model that has been helpful in finding funding for RSE groups is the use of
funds left over on research grants when RAs have left prematurely -- PIs like
this arrangement as it is hard to find good staff for short-term positions, so
having a pool of research programming staff on hand resolves this problem. We
recommend that funders give explicit guidance to grant holders and institutions
that such arrangement are favorable. Framework agreements permitting this to go
ahead without checking back every time with funders and/or grant panels would
further smooth this. (This also provides more stable jobs for those who hold
the skills, but arguments about making life nicer for postdocs will not help
persuade the funders or the PIs.)

There is some questions about the most effective duration and percent of full
time for a programmer's work on a project. At least three months is necessary
for the programmer to read into the science (RSEs must not become so disengaged
from research that they do not have time to read a few papers -- this will result
in code which does not meet scientific needs), but too long could result in an
RSE losing their flexibility, becoming so engaged in one project that when that
project ends, they find it hard to transfer. For this reason, we also recommend
that 40 percent is ideal; two projects per developer, with some time for
training and infrastructure work. Two developers per project seems to be ideal,
in the sense that software development is enhanced by two pairs of eyes.

There is not yet clear answer to the scale of aggregation needed to make
such a program work. A university wide program allows enough scale to be robust
to fluctuations of funding within one field. But a specialization focus on
developers to support, for example, physical or biological sciences may be
preferable, if the customer base is large enough. The desire to aggregate enough
work to make it sustainable, and the need to have domain-relevant research
programming skills, are in tension.

@@

However this area develops, the need for funding to secure software as a
cyberinfrastructure component is clear. \choinote{last sentence unclear} Funding which permits code to be
refactored, tidied and optimized is rare; this is often done ``on the sly'' \choinote{on the fly?} in a
scientifically focused grant. The UK EPSRC's ``software for the future'' call,
which really permits explicit investment in software as an infrastructure, is so
oversubscribed as to have a 4\% success rate; the demand is clear!

@@

It can be a more difficult to help leading domain scientists see the value of
engineering effort than those in their teams who are forced to work with
difficult-to-use or unreliable software tools, as they do not see the pain.
Perhaps a version of ``software carpentry'' targeted at those PIs who are
awarded or apply for software grants could be valuable here.

@@

We will seek to identify and approach existing research programming organizations,
to get their permission to list them on a list of research software groups.
Casual conversation during the meeting made it clear that although the title is
not widely used in the US,  this position is not rare. We spoke with several
individuals who, at distinct universities, had RSEs (in effect if not in name)
and these were funded under differing models.

@@

In this respect, information gathering via a survey and subsequent analysis could be
very useful. We would need to assemble a list of targeted individuals (what
positions and ranks are likely to know and care enough to respond?). Perhaps the
Science Gateway Institute has already acquired information that could be helpful
to advance this issue, and/or craft a proper survey and target some individuals.

>>>>

What is the power of the label? How can we get the word out about RSE support using the label?

@@

An alternative approach has been the provision of programming expertise as part
of national compute services. The US XSEDE project's Extended Collaboration Support 
Services (ECSS) is a set of developers who are paid with XSEDE funding, and are
on ``permanent'' staff. When PIs request allocations on XSEDE resources, there
is a finite pool of developer time that can be awarded, typically for one year
only, and at partial effort, typically 20 percent or so. The finite time allowed
provides motivation for the scientist and the scientist's group to work closely with the
developer and to become educated in what the developer is doing, so they can
sustain the effort once the ECSS period is over. This funding mechanism can be
highly efficient for scientific problems, because the developer pool assembled
by the research providers are, by definition, expert in the characteristics of
their specific resource, and can very quickly assess the scientist's needs, and
what it will take to implement software that meets the user's needs. However, it
does not develop capacity within institutions, and since XSEDE is a time-bounded
program, it should not be relied upon as a long-term solution to acquiring this
type of capacity.

The UK allows this kind of collaboration to support the creation of scientific
software for the large supercomputing resource (ARCHER). However, while the
support can come the staff of the Edinburgh Parallel Computing Centre,
who host the computer, this ``embedded CSE'' resource also funds the
programming coming from local groups. This has been very helpful in providing
funding to establish local groups. These work best when these groups develop good
collaborations with national cyberinfrastructure pools. When an organization
assembles a developer pool, diversity is developed and skills can be
transferred.

@@

The CANARIE group (Canada) accepts proposals for providing services to broad
communities, integrating people who are doing things that are complementary. The
goal is to make the available stack more robust and richer for everyone. They
offer short cycles of funding that can have use creating some useful
functionality that shows a diversity of input and draws from across disciplines as a key metric,  If this metric is met successfully, 
then more funding may follow. This can apply within or across institutions.

There can be problems communicating across cultural barriers, with domain
scientists seeing developers as ``other''. Both collaboration and tools to fund,
encourage, or motivate collaboration are extremely important.

We think support from non-governmental organizations will be important if RSE
groups will become established. The Sloan Foundation is currently funding data
science engineers, who work in the context of other software developers at the University
of Washington. These scientists work in the e-Science Studio/Data Science
Studio, and they help a group of graduate students in solving their problems in
data science and data management. During Fall and Spring, a 10-week incubator
program allows students to work two days a week on a data-intensive
science project. Some fraction of the developer time is dedicated to the
developers' personal interests as well as instruction.

The goal for Sloan is to obtain success stories and demonstrable value
of the presence of data scientists on university staff. These stories are the
basis for arguments to the host organization. This is an effort to create
awareness of the value of research scientist developers. Embedding with
scientists, and adding spare capacity is critical to making the innovation
possible. This model is essentially to argue for permanent budget lines to
support data scientists as part of university staff hires, just as with core
facilities. This could become a fee-for-service model requested by grant
funding, just as DNA sequencing is for core facilities, if it becomes apparent
that this gives competitive advantage to a university's research effort.

One model that has been helpful in finding funding for RSE groups is the use of
funds left over on research grants when RAs have left prematurely -- PIs like
this arrangement as it is hard to find good staff for short-term positions, so
having a pool of research programming staff on hand resolves this problem. We
recommend that funders give explicit guidance to grant holders and institutions
that such arrangement are favorable. Framework agreements permitting this to go
ahead without checking back every time with funders and/or grant panels would
further smooth this. (This also provides more stable jobs for those who hold
these skills, but arguments about making life nicer for postdocs will not help
persuade funders or PIs!)

There is some question about the most effective duration and percent of full
time for a programmer's work on a project. At least three months is necessary
for the programmer to read into the science (RSEs must not become so disengaged
from research that they do not have time to read a few papers -- this will result
in code which does not meet scientific needs), but too long could result in an
RSE losing their flexibility, becoming so engaged in one project that when that
project ends, they find it hard to transfer. For this reason, we also recommend
that 40 percent is ideal; two projects per developer, with some time for
training and infrastructure work. Two developers per project seems to be ideal,
in the sense that software development is enhanced by two pairs of eyes.

There is, as yet, no clear answer as to the scale of aggregation needed to make
such a program work. A university wide program allows enough scale to be robust
to fluctuations of funding within one field. But a specialization focus on
developers to support, for example, physical or biological sciences may be
preferable, if the customer base is large enough. The desire to aggregate enough
work to make it sustainable, and the need to have domain-relevant research
programming skills, are in tension.

@@

However this area develops, the need for funding for software as a
cyberinfrastructure component is clear.  Funding that permits code to be
refactored, tidied, and optimized is rare; this is often done ``on the sly'' in a
scientifically focused grant. The UK EPSRC's ``software for the future'' call,
which really permits explicit investment in software as an infrastructure, is so
oversubscribed as to have a 4\% success rate; the demand is clear!

@@

It can be a more difficult to help leading domain scientists see the value of
engineering effort than those in their teams who are forced to work with
difficult-to-use or unreliable software tools, as they do not see the pain.
Perhaps a version of ``software carpentry'' targeted at those PIs who are
awarded or apply for software-intensive grants could be valuable here.

@@

We will seek to identify and approach existing research programming organizations,
to get their permission to list them on a list of research software groups.
Casual conversation during the meeting made it clear that although the title is
not widely used in the US, this position is not rare. We spoke with several
individuals who, at distinct universities, had RSEs (in effect if not in name)
who were funded under differing models.

@@

In this respect, information gathering via a survey and subsequent analysis could be
very useful. We would need to assemble a list of targeted individuals. (what
positions and ranks are likely to know and care enough to respond?) Perhaps the
Science Gateway Institute has already acquired information that could be helpful
to advance this issue, and/or craft a proper survey and suggest target individuals.
